---
title: "Uber Data Analysis in R"
author: "Ebenezer Bediam Nonbo"
date: "2023-04-29"
output: html_document
---



Uber Data Analysis through Visualisations in R

Data storytelling is an important component of Machine Learning through which companies are able to understand the background of various operations. With the help of visualization, companies can avail the benefit of understanding the complex data and gain insights that would help them to craft decisions. This is more of a data visualization project that will guide you towards using the ggplot2 library for understanding the data and for developing an intuition for understanding the customers who avail the trips.

Important: The goal of this project is to learn visualisations in R


```{r}
library(ggplot2)
library(ggthemes)
library(lubridate)
library(dplyr)
library(tidyr)
library(tidyverse) # metapackage of all tidyverse packages
library(DT)         #  Datatables in JS
library(scales)   # map the data to the correct scales with well-placed axes and legends.

```

## 

```{r}
## Creating vector of colors for the plots
colors = c("#CC1011", "#665555", "#05a399", "#cfcaca", "#f5e840", "#0683c9", "#e075b0")
colors

```

## 

```{r}
## Read the data from each time-frame
# Read the data for each month separately 
apr <- read.csv("D:/BEDIAM/Ebenezer/Logiciels/R/Machine_learning/archive/uber-raw-data-apr14.csv")
may <- read.csv("D:/BEDIAM/Ebenezer/Logiciels/R/Machine_learning/archive/uber-raw-data-may14.csv")
june <- read.csv("D:/BEDIAM/Ebenezer/Logiciels/R/Machine_learning/archive/uber-raw-data-jun14.csv")
july <- read.csv("D:/BEDIAM/Ebenezer/Logiciels/R/Machine_learning/archive/uber-raw-data-jul14.csv")
aug <- read.csv("D:/BEDIAM/Ebenezer/Logiciels/R/Machine_learning/archive/uber-raw-data-aug14.csv")
sept <- read.csv("D:/BEDIAM/Ebenezer/Logiciels/R/Machine_learning/archive/uber-raw-data-sep14.csv")

# Combine the data together 
data <- rbind(apr, may, june, july, aug, sept)
cat("The dimensions of the data are:", dim(data))

```

```{r}
##The data contains the columns Date.Time which is a factor, Latitude and Longitudes which are double and Base which is factor. we will format the datetime into a more readable format using the Date Time conversion function.

data$Date.Time <- as.POSIXct(data$Date.Time, format="%m/%d/%Y %H:%M:%S")
data$Time <- format(as.POSIXct(data$Date.Time, format = "%m/%d/%Y %H:%M:%S"), format="%H:%M:%S")
data$Date.Time <- ymd_hms(data$Date.Time)

```

## 

```{r}
# Create individual columns for month day and year
data$day <- factor(day(data$Date.Time))
data$month <- factor(month(data$Date.Time, label=TRUE))
data$year <- factor(year(data$Date.Time))
data$dayofweek <- factor(wday(data$Date.Time, label=TRUE))

```

## 

```{r}
# Add Time variables as well 
data$second = factor(second(hms(data$Time)))
data$minute = factor(minute(hms(data$Time)))
data$hour = factor(hour(hms(data$Time)))

```


##  Data Visualisation

```{r}
## Plotting the trips by hours in a day
hourly_data <- data %>% 
                    group_by(hour) %>% 
                            dplyr::summarize(Total = n())

# Shos data in a searchable js table
datatable(hourly_data)

```

## 

```{r}
# Plot the data by hour
ggplot(hourly_data, aes(hour, Total)) + 
geom_bar(stat="identity", 
         fill="steelblue", 
         color="red") + 
ggtitle("Trips Every Hour", subtitle = "aggregated today") + 
theme(legend.position = "none", 
      plot.title = element_text(hjust = 0.5), 
      plot.subtitle = element_text(hjust = 0.5)) + 
scale_y_continuous(labels=comma)

```



```{r}
## Plotting trips by hour and month
# Aggregate the data by month and hour
month_hour_data <- data %>% group_by(month, hour) %>%  dplyr::summarize(Total = n())

ggplot(month_hour_data, aes(hour, Total, fill=month)) + 
geom_bar(stat = "identity") + 
ggtitle("Trips by Hour and Month") + 
scale_y_continuous(labels = comma)
```


```{r}
## Plotting data by trips during every day of the month
# Aggregate data by day of the month 
day_data <- data %>% group_by(day) %>% dplyr::summarize(Trips = n())
day_data

```

## 

```{r}
# Plot the data for the day
ggplot(day_data, aes(day, Trips)) + 
geom_bar(stat = "identity", fill = "steelblue") +
ggtitle("Trips by day of the month") + 
theme(legend.position = "none") + 
scale_y_continuous(labels = comma)

```

## 

```{r}
# Collect data by day of the week and month

day_month_data <- data %>% group_by(dayofweek, month) %>% dplyr::summarize(Trips = n())
day_month_data

```


```{r}
# Plot the above data
ggplot(day_month_data, aes(dayofweek, Trips, fill = month)) + 
geom_bar(stat = "identity", aes(fill = month), position = "dodge") + 
ggtitle("Trias by Day and Month") + 
scale_y_continuous(labels = comma) + 
scale_fill_manual(values = colors)

```

## 

```{r}
##  Number of Trips place during months in a year


month_data <- data %>% group_by(month) %>% dplyr::summarize(Total = n())

month_data

```

## 

```{r}
ggplot(month_data, aes(month, Total, fill = month)) + 
geom_bar(stat = "Identity") + 
ggtitle("Trips in a month") + 
theme(legend.position = "none") + 
scale_y_continuous(labels = comma) + 
scale_fill_manual(values = colors)

```


## Heatmap visualization of day, hour and month

```{r}
## Heatmap by Hour and Day
day_hour_data <- data %>% group_by(day, hour) %>% dplyr::summarize(Total = n())
datatable(day_hour_data)

```

## 

```{r}
# Plot a heatmap 

ggplot(day_hour_data, aes(day, hour, fill = Total)) + 
geom_tile(color = "white") + 
ggtitle("Heat Map by Hour and Day")

```

## 

```{r}
## Plot Heatmap by day and month
# Collect data by month and day

month_day_data <- data %>% group_by(month, day) %>% dplyr::summarize(Trips = n())
month_day_data

```


```{r}
# Plot a heatmap 

ggplot(month_day_data, aes(day, month, fill = Trips)) + 
geom_tile(color = "white") + 
ggtitle("Heat Map by Month and Day")

```

## 

```{r}
# Plot a heatmap by day of the week and month

ggplot(day_month_data, aes(dayofweek, month, fill = Trips)) + 
geom_tile(color = "white") + 
ggtitle("Heat Map by Month and Day")

```

## 

```{r}
## Creating a map visualization of rides in NYC
# Set Map Constants
min_lat <- 40 
max_lat <- 40.91
min_long <- -74.15
max_long <- -73.7004

```


```{r}
ggplot(data, aes(x=Lon, y=Lat)) +
  geom_point(size=1, color = "blue") +
     scale_x_continuous(limits=c(min_long, max_long)) +
      scale_y_continuous(limits=c(min_lat, max_lat)) +
        theme_map() +
           ggtitle("NYC MAP BASED ON UBER RIDES DURING 2014 (APR-SEP)")

```

## 

```{r}

```

## 

```{r}

```

