---
title: "Data Exploration with RMS Titanic"
author: "Ebenezer Bediam Nonbo"
date: "2023-04-03"
output: html_document
---



## Data Exploration with RMS Titanic


```{r}
#In a real data exploration project, there are six steps involved in the exploration process. They are as follows: Asking the right questions,Data collection,Data munging,Basic exploratory data analysis,Advanced exploratory data analysis,Model assessment.

library(readxl)
tested_xls<-read_excel("D:/BEDIAM/Ebenezer/Logiciels/R/Machine_learning/tested.xls.xlsx")
View(tested_xls)
str(tested_xls)

# NB:In R,different measurements are calculated differently, so you should perform a type conversion before applying descriptive or inferential analytics toward the dataset.
#Convert that column to a factor
 tested_xls$ Pclass <- factor(tested_xls$ Pclass)
 tested_xls$ Survived <- factor(tested_xls$ Survived)
 str(tested_xls)

```

## 

```{r}
## Detecting missing values

#The is.na function is used to denote which index of the attribute contains the NA. we apply it to the Age attribute first:
 is.na(tested_xls$Age)
#To get a general number of how many missing values there are
 sum(is.na(tested_xls$Age) == TRUE)
#To calculate the percentage of missing values
 sum(is.na(tested_xls$Age) == TRUE) / length(tested_xls$Age)
#To get a percentage of the missing value of the attributes
 sapply(tested_xls, function(df) { + sum(is.na(df)==TRUE)/ length(df) } )
#use the Amelia package to visualize the missing values 
 #install.packages("Amelia")
 library(Amelia)
#Then, use the missmap function to plot the missing value map
 missmap(tested_xls, main="Missing Map")
#To start running AmeliaView, simply type AmeliaView() in the R Console

```

## 

```{r}
##Imputing missing values

#Perform the following steps to impute the missing values
#First, list the distribution of Port of Embarkation. Here, we add the useNA = "always" argument to show the number of NA values
 table(tested_xls$Embarked, useNA = "always")
#Assign the missing values to a more probable port (that is, the most counted port)
 tested_xls$Embarked[which(is.na(tested_xls$Embarked))] = 'S';
 table(tested_xls$Embarked, useNA = "always")
#In order to discover the types of titles contained in the names of train.data
 tested_xls$Name = as.character(tested_xls$Name)
#we first tokenize train.data$Name by blank (a regular expression pattern as "\\s+")
#then count the frequency of occurrence with the table function
 table_words = table(unlist(strsplit(tested_xls$Name, "\\s+")))
#we use the regular expression to grep the word containing the period
 sort(table_words [grep('\\.',names(table_words))], decreasing=TRUE)
#To obtain which title contains missing values, you can use str_match provided by the stringr package to get a substring containing a period
#bind the column together with cbind
 library(stringr) 
 tb = cbind(tested_xls$Age, str_match(tested_xls$Name, " [a-zA-Z]+\\."))
#using the table function to acquire the statistics of missing values
 table(tb[is.na(tb[,1]),2])
#Assign the mean value for each title (not containing a missing value)
 tested_xls$Age <- as.numeric(tested_xls$Age)
  mean.mr = mean(tested_xls$Age[grepl(" Mr\\.", tested_xls$Name) & !is.na(tested_xls$Age)])
  mean.mrs = mean(tested_xls$Age[grepl(" Mrs\\.", tested_xls$Name) & !is.na(tested_xls$Age)])
 mean.dr = mean(tested_xls$Age[grepl(" Dr\\.", tested_xls$Name) & !is.na(tested_xls$Age)])
 mean.miss = mean(tested_xls$Age[grepl(" Miss\\.", tested_xls$Name) & !is.na(tested_xls$Age)])
 mean.master = mean(tested_xls$Age[grepl(" Master\\.", tested_xls$Name) & !is.na(tested_xls$Age)])
#Then, assign the missing value with the mean value of each title:
 tested_xls$Age[grepl(" Mr\\.", tested_xls$Name) & is.na(tested_xls$Age)] = mean.mr
 tested_xls$Age[grepl(" Mrs\\.", tested_xls$Name) & is.na(tested_xls$Age)] = mean.mrs 
 tested_xls$Age[grepl(" Dr\\.", tested_xls$Name) & is.na(tested_xls$Age)] = mean.dr
 tested_xls$Age[grepl(" Miss\\.", tested_xls$Name) & is.na(tested_xls$Age)] = mean.miss 
 tested_xls$Age[grepl(" Master\\.", tested_xls$Name) & is.na(tested_xls$Age)] = mean.master 

```

## 

```{r}
## Exploring and visualizing data

#First, you can use a bar plot and histogram to generate descriptive statistics for each attribute starting with passenger survival:
 barplot(table(tested_xls$Survived), main="Passenger Survival", names= c("Perished", "Survived"))
#We can generate the bar plot of passenger class
 barplot(table(tested_xls$Pclass), main="Passenger Class", names= c("first", "second", "third"))
#Next, we outline the gender data with the bar plot
 barplot(table(tested_xls$Sex), main="Passenger Gender")
#We then plot the histogram of the different ages with the hist function
 hist(tested_xls$Age, main="Passenger Age", xlab = "Age")
#We can plot the bar plot of sibling passengers to get the following
 barplot(table(tested_xls$SibSp), main="Passenger Siblings")
#Next, we can get the distribution of the passenger parch
 barplot(table(tested_xls$Parch), main="Passenger Parch")
#Next, we plot the histogram of the passenger fares
 hist(tested_xls$Fare, main="Passenger Fare", xlab = "Fare")
#Finally, one can look at the port of embarkation
 barplot(table(tested_xls$Embarked), main="Port of Embarkation")
#Use barplot to find out which gender is more likely to perish during shipwrecks
 counts = table( tested_xls$Survived, tested_xls$Sex)
 barplot(counts, col=c("darkblue","red"), legend = c("Perished", "Survived"), main = "Passenger Survival by Sex")
#Next, we should examine whether the Pclass factor of each passenger may affect the survival rate 
 counts = table( tested_xls$Survived, tested_xls$Pclass)
 barplot(counts, col=c("darkblue","red"), legend =c("Perished", "Survived"), main= "Titanic Class Bar Plot" )  
#Next, we examine the gender composition of each Pclass
 counts = table( tested_xls$Sex, tested_xls$Pclass)
 barplot(counts, col=c("darkblue","red"), legend = rownames(counts), main= "Passenger Gender by Class")  
#Furthermore, we examine the histogram of passenger ages (Launch both graphs at the same time) 
 hist(tested_xls$Age[which(tested_xls$Survived == "0")], main= "Passenger Age Histogram", xlab="Age", ylab="Count", col ="blue", breaks=seq(0,80,by=2))
 hist(tested_xls$Age[which(tested_xls$Survived == "1")], col ="red", add = T, breaks=seq(0,80,by=2)) 
#To examine more details about the relationship between the age and survival rate, one can use a boxplot  
boxplot(tested_xls$Age ~ tested_xls$Survived) 
#To categorize people with different ages into different groups, such as children (below 13), youths (13 to 19), adults (20 to 65), and senior citizens (above 65), execute the following commands
 train.child = tested_xls$Survived[tested_xls$Age < 13]
 length(train.child[which(train.child == 1)] ) / length(train.child)
 
 train.youth = tested_xls$Survived[tested_xls$Age >= 15 & tested_xls$Age < 25]
 length(train.youth[which(train.youth == 1)] ) / length(train.youth)
 
 train.adult = tested_xls$Survived[tested_xls$Age >= 20 & tested_xls$Age < 65]
 length(train.adult[which(train.adult == 1)] ) / length(train.adult)
 
 train.senior = tested_xls$Survived[tested_xls$Age >= 65]
 length(train.senior[which(train.senior == 1)] ) / length(train.senior)
 #one can also apply mosaicplot in the vcd package to examine the relationship between multiple categorical variables
  mosaicplot(tested_xls$Pclass ~ tested_xls$Survived,color=TRUE) 

```

## 

```{r}
## Predicting passenger survival with a decision tree
  
#The exploratory analysis helps users gain insights into how single or multiple variables may affect the survival rate. However, it does not determine what combinations may generate a prediction model, so as to predict the passengers' survival 
#On the other hand, machine learning can generate a prediction model from a training dataset 
 tested_xls$Sex <- as.factor(tested_xls$Sex)
 tested_xls$Fare <- as.factor(tested_xls$Fare)
 tested_xls$Embarked <- as.factor(tested_xls$Embarked)
# Data Partition:
 set.seed(123)
 ind <- sample(2,nrow(tested_xls),replace = TRUE,prob = c(0.7,0.3))
 train <- tested_xls[ind==1,]
 test <- tested_xls[ind==2,] 
#For the condition tree, one has to use the ctree function from the party package; therefore, we install and load the party package:  
 install.packages('party')
 library(party)
#We then use Survived as a label to generate the prediction model in use. After that, we assign the classification tree model into the train.ctree variable 
 train.ctree = ctree(Survived ~ Pclass + Sex + Age + SibSp + Fare + Parch + Embarked, data=train)
 train.ctree 
#We use a plot function to plot the tree
 plot(train.ctree, main="Conditional inference tree of Titanic Dataset") 
#Besides ctree, one can also use svm to generate a prediction model. To load the svm 
 install.packages('e1071')
 library(e1071)
 svm.model = svm(Survived ~ Pclass + Sex + Age + SibSp + Fare + Parch + Embarked, data = train, probability = TRUE)
 svm.model 

```

## 

```{r}
## Validating the power of prediction with a confusion matrix
 
#For now, users will learn to validate how well ctree performs in a survival prediction via the use of a confusion matrix.
#We start using the constructed train.ctree model to predict the survival of the testing set 
 ctree.predict = predict(train.ctree, test) 
#First, we install the caret package, and then load it:
 install.packages("caret")
 library(caret) 
#After loading caret, one can use a confusion matrix to generate the statistics of the output matrix:
confusionMatrix(ctree.predict, test$Survived) 
#NB:One may generate a better prediction model by tuning the attribute used, or by replacing the classification algorithm to SVM, glm, or random forest.

```

```

## 

```{r}
## Assessing performance with the ROC curve

#Another measurement is by using the ROC curve (this requires the ROCR package)
#which plots a curve according to its true positive rate against its false positive rate
#be sure that the generated training set, testing dataset, and built prediction model, ctree.predict, are within the R session
#Prepare the probability matrix:
 train.ctree.pred = predict(train.ctree, test)
 train.ctree.prob = 1- unlist(treeresponse(train.ctree, test), use.names=F)[seq(1,nrow(test)*2,2)]
#Install and load the ROCR package:
 install.packages("ROCR")
 library(ROCR)
#Create an ROCR prediction object from probabilities
 train.ctree.prob.rocr = prediction(train.ctree.prob,test$Survived)
#Prepare the ROCR performance object for the ROC curve (tpr=true positive rate, fpr=false positive rate) and the area under curve (AUC)
 train.ctree.perf = performance(train.ctree.prob.rocr, "tpr","fpr")
 train.ctree.auc.perf = performance(train.ctree.prob.rocr, measure = "auc", x.measure = "cutoff")
#Plot the ROC curve, with colorize as TRUE, and put AUC as the title: 
  plot(train.ctree.perf, col=2,colorize=T, main=paste("AUC:", train.ctree.auc.perf@y.values))
  
```

## 
